A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-10-26 imaging_demo/ms.tex
fixing more typos

We provide a brief summary of the amthematical equivalence of the MOFF and 
FX correlators detailed in \citet{mor11}. We first relate the dirty image produced 
from visibilities to the electric fields of astrophysical sources, then show that 
operations can be reordered to produce the same images at a lower 
computational cost.

@@

Interferometers measure {\it visibilities} -- the degree of coherence between
electric fields measured by a pair of antennas \citep{van34,zer38,tho01}.
A visibility, $\widetilde{V}_p$, can be written as:
\begin{align}
  \widetilde{V}_p = &\left\langle \widetilde{E}_a\widetilde{E}_b^\star \right\rangle_t \label{eqn:cc-vis}\hfill\\
  		= &\left\langle \left[ \int {W}_a(\hat{\mathbf{s}})\,E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_a \right] \right. \nonumber\\
		&\qquad \times \left.\left[ \int {W}^\star_b(\hat{\mathbf{s'}})\,E^\star (\hat{\mathbf{s'}})\,e^{i2\pi\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}}\,\dif^2\hat{\mathbf{s'}} + \widetilde{n}^\star_b \right] \right\rangle_t \\
                  =& \iint {W}_a(\hat{\mathbf{s}}) {W}^\star_b(\hat{\mathbf{s'}}) \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) \right\rangle_t e^{-i2\pi(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})}\,\dif^2\hat{\mathbf{s}}\,\dif^2\hat{\mathbf{s'}},
\end{align}
where we have brought the time average into the integral under the assumption that 
the aperture illumination pattern does not change over the time-scale of the 
averaging. This expression can be further simplified with the sky brightness, 
$I(\hat{\mathbf{s}})= \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) 
\right\rangle_t  \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the 
antenna pair sky power response function (or the primary beam), 
$B_p(\hat{\mathbf{s}})~\equiv~{W}_a(\hat{\mathbf{s}})~{W}^\star_b(\hat{\mathbf{s}})$. 
The result is the visibility expressed in terms of the sky brightness, the primary 
beam, and uncorrelated noise terms which we group into $\widetilde{n}_p$,
\begin{equation}
\widetilde{V}_p = \int e^{-i2\pi\mathbf{u}_p\!\cdot\,\hat{\mathbf{s}}}\,B_p(\hat{\mathbf{s}})\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_p,
\end{equation}
where the baseline coordinate $\mathbf{u}_p=\mathbf{r}_a-\mathbf{r}_b$ is the 
vector separation between the two antennas. This signifies that the visibility 
($\widetilde{V}_p$) measured between a
pair of antennas ($p$) is obtained by the multiplying the sky brightness
$I(\hat{\mathbf{s}})$ by the antenna power response $B(\hat{\mathbf{s}})$ and
Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to $uv$
coordinates, which are then sampled at the locations of the antenna spacings (or
baselines), namely, $\mathbf{u}_p$, and added to the receiver noise $n_p$.

@@

From here we adopt the matrix notation of \citet{mor11}, where vectors are 
represented with single coordinates, and matrices are represented by two 
coordinates denoting the spaces the operator transforms between. 
In this notation, the above measurement equation can be expressed as:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \widetilde{\mathbf{B}}(\mathbf{v},\mathbf{u})\,\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})\,\mathbf{I}(\hat{\mathbf{s}}) + \mathbf{n}(\mathbf{v}),
\end{align}
where the sky brightness $\mathbf{I}(\hat{\mathbf{s}})$ is Fourier transformed using
$\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})$ and the resultant spatial coherence
function is weighted and summed using the antenna power response,
$\widetilde{\mathbf{B}}(\mathbf{v},\mathbf{u})$ in $uv$-space sampled at the baseline
location to obtain the measured visibilities:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \left\langle \widetilde{\mathbf{E}}^\star(\mathbf{a})\,\widetilde{\mathbf{E}}(\mathbf{a}^\prime)\right\rangle_t, \label{eqn:matrix-cc-vis}
\end{align}
where $\mathbf{m}(\mathbf{v})$ denotes visibilities measured by cross-correlating
measured antenna electric fields over all possible pairs of $\mathbf{a}$ and
$\mathbf{a}^\prime$. It is the same as equation~\ref{eqn:cc-vis} written in
matrix notation.

@@

We have implemented the MOFF imaging technique in our ``E-field Parallel Imaging
Correlator'' -- a highly parallelized Object Oriented Python package,
\footnote{EPIC package can be accessed at https://github.com/nithyanandan/EPIC}
now publicly available. Besides implementing the MOFF imaging algorithm it also
includes FX imaging using the software holography technique and a simulator for
generating electric fields from a sky model.

@@

In a companion paper to appear soon, we demonstrate a novel calibration 
technique (EPICal) which leverages the data products formed by direct imaging 
correlators to estimate antenna complex gains. This method correlates the antenna 
electric field signals with an image pixel form the output of the correlator in 
the feedback calibration fashion outlined in \citealt{mor11} (illustrated in Fig. 
\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the calibration 
block). Furthermore it allows for arbitrarily complex sky models, and following 
the MOFF algorithm places no restriction on array layout, and accounts for 
non-identical antenna beam patterns. Because only a single correlation is needed 
for each antenna, the computation complexity scales only as $N_{\mathrm{ant}}$.

@@

The Antenna module is a fundamental building block upon which all the other 
modules are built. There is one Antenna module per antenna each having 
attributes -- the propagated electric field time-series, $E(t)$, and spectrum 
$E(f)$ for both polarizations. The most important function inside this module 
is the F-engine that Fourier transforms time-series electric field data into 
spectrum.

@@

Fig.~\ref{fig:LWA-image} shows the image produced with MOFF imaging 
packaged in EPIC after averaging over the entire 2~s of data and the inner
$\approx 80$\% of bandwidth (roughly 80~kHz). The image is shown in direction 
cosine coordinates -- $l$ along x-axis and $m$ along y-axis. The flux scale 
is arbitrary. Even in this proof-of-concept demonstration, we see Cyg~A and 
Cas~A prominently and Sgr~A faintly as annotated, thus validating the 
functionality of the EPIC package.

>>>>

We provide a brief summary of the mathematical equivalence of the MOFF and 
FX correlators detailed in \citet{mor11}. We first relate the dirty image produced 
from visibilities to the electric fields of astrophysical sources, then show that 
operations can be reordered to produce the same images at a lower 
computational cost.

@@

Interferometers measure {\it visibilities} -- the degree of coherence between
electric fields measured by a pair of antennas \citep{van34,zer38,tho01}.
A visibility, $\widetilde{V}_p$, can be written as:
\begin{align}
  \widetilde{V}_p = &\left\langle \widetilde{E}_a\widetilde{E}_b^\star \right\rangle_t \label{eqn:cc-vis}\hfill\\
  		= &\left\langle \left[ \int {W}_a(\hat{\mathbf{s}})\,E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_a \right] \right. \nonumber\\
		&\qquad \times \left.\left[ \int {W}^\star_b(\hat{\mathbf{s'}})\,E^\star (\hat{\mathbf{s'}})\,e^{i2\pi\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}}\,\dif^2\hat{\mathbf{s'}} + \widetilde{n}^\star_b \right] \right\rangle_t \\
                  =& \iint {W}_a(\hat{\mathbf{s}}) {W}^\star_b(\hat{\mathbf{s'}}) \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) \right\rangle_t e^{-i2\pi(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})}\,\dif^2\hat{\mathbf{s}}\,\dif^2\hat{\mathbf{s'}},
\end{align}
where we have brought the time average into the integral under the assumption that 
the aperture illumination pattern does not change over the time-scale of the 
averaging. This expression can be further simplified with the sky brightness, 
$I(\hat{\mathbf{s}})= \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) 
\right\rangle_t  \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the 
antenna pair sky power response function (or the primary beam), 
$B_p(\hat{\mathbf{s}})~\equiv~{W}_a(\hat{\mathbf{s}})~{W}^\star_b(\hat{\mathbf{s}})$. 
The result is the visibility expressed in terms of the sky brightness, the primary 
beam, and uncorrelated noise terms which we group into $\widetilde{n}_p$,
\begin{equation}
\widetilde{V}_p = \int e^{-i2\pi\mathbf{u}_p\!\cdot\,\hat{\mathbf{s}}}\,B_p(\hat{\mathbf{s}})\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_p,
\end{equation}
where the baseline coordinate $\mathbf{u}_p=\mathbf{r}_a-\mathbf{r}_b$ is the 
vector separation between the two antennas. This signifies that the visibility 
($\widetilde{V}_p$) measured between a
pair of antennas ($p$) is obtained by the multiplying the sky brightness
$I(\hat{\mathbf{s}})$ by the antenna power response $B_p(\hat{\mathbf{s}})$ and
Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to $uv$
coordinates, which are then sampled at the locations of the antenna spacings (or
baselines), namely, $\mathbf{u}_p$, and added to the receiver noise $n_p$.

@@

From here we adopt the matrix notation of \citet{mor11}, where vectors are 
represented with single coordinates, and matrices are represented by two 
coordinates denoting the spaces the operator transforms between. 
In this notation, the above measurement equation can be expressed as:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \widetilde{\mathbf{B}}(\mathbf{v},\mathbf{u})\,\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})\,\mathbf{I}(\hat{\mathbf{s}}) + \mathbf{n}(\mathbf{v}),
\end{align}
where the sky brightness $\mathbf{I}(\hat{\mathbf{s}})$ is Fourier transformed using
$\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})$ and the resultant spatial coherence
function is weighted and summed using the antenna power response,
$\widetilde{\mathbf{B}}(\mathbf{v},\mathbf{u})$ in $uv$-space sampled at the baseline
location to obtain the measured visibilities:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \left\langle \widetilde{\mathbf{E}}(\mathbf{a})\,\widetilde{\mathbf{E}}^\star(\mathbf{a}^\prime)\right\rangle_t, \label{eqn:matrix-cc-vis}
\end{align}
where $\mathbf{m}(\mathbf{v})$ denotes visibilities measured by cross-correlating
measured antenna electric fields over all possible pairs of $\mathbf{a}$ and
$\mathbf{a}^\prime$. It is the same as equation~\ref{eqn:cc-vis} written in
matrix notation.

@@

We have implemented the MOFF imaging technique in our ``E-field Parallel Imaging
Correlator'' -- a highly parallelized Object Oriented Python 
package,\footnote{EPIC package can be accessed at https://github.com/nithyanandan/EPIC}
now publicly available. Besides implementing the MOFF imaging algorithm it also
includes FX imaging using the software holography technique and a simulator for
generating electric fields from a sky model.

@@

In a companion paper to appear soon, we demonstrate a novel calibration 
technique (EPICal) which leverages the data products formed by direct imaging 
correlators to estimate antenna complex gains. This method correlates the antenna 
electric field signals with an image pixel from the output of the correlator in 
the feedback calibration fashion outlined in \citealt{mor11} (illustrated in Fig. 
\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the calibration 
block). Furthermore it allows for arbitrarily complex sky models, and following 
the MOFF algorithm places no restriction on array layout, and accounts for 
non-identical antenna beam patterns. Because only a single correlation is needed 
for each antenna, the computation complexity scales only as $N_{\mathrm{a}}$.

@@

The Antenna module is a fundamental building block upon which all the other 
modules are built. There is one Antenna module per antenna each having 
attributes -- the propagated electric field time-series, $E(t)$, and spectrum 
$E(f)$ for both polarizations. The most important function inside this module 
is the F-engine that Fourier transforms time-series electric field data into 
spectra.

@@

Fig.~\ref{fig:LWA-image} shows the image produced with MOFF imaging 
packaged in EPIC after averaging over the entire 2~s of data and the inner
$\approx 80$\% of bandwidth (roughly 80~kHz). The image is shown in direction 
cosine coordinates -- $l$ along the horizontal axis and $m$ along the vertical axis. The flux scale 
is arbitrary. Even in this proof-of-concept demonstration, we see Cyg~A and 
Cas~A prominently and Sgr~A faintly as annotated, thus validating the 
functionality of the EPIC package.
